---
title: "Reproducible Research, Week 2, Course Project 1"
author: "Data Analyst: Michael A. Esparza, PMP, PgMP, CPEM"
date: "Date Submitted: June 10, 2018"
output: html_document  

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Git Hub URL: https://github.com/MichaelEsparza012/RepData_PeerAssessment1   
## Specfied Tasks for this Assignment:  
### Loading and preprocessing the data  
#### 1. Show any code that is needed to Load the data 
#### 2. Process/transform the data (if necessary) into a format suitable for your analysis  
##### Download the data frame from the course website: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip  

```{r}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", destfile = "activity.zip", mode="wb")
```
#### Unzip the datafiles, extract and read for confirmation
```{r}
unzip("activity.zip")
stepdata <- read.csv("activity.csv", header = TRUE)
head(stepdata)
```
### What is mean total number of steps taken per day?  
#### 1. Calculate the total number of steps taken per day  
```{r}
library(magrittr)
library(dplyr)
```
#### 2. Make a histogram of the total number of steps taken each day (a Bar Chart assigns Categorical Values, a Histogram assigns Quantitative Values) 
```{r}
databydate <- stepdata %>% select(date, steps) %>% group_by(date) %>% summarize(tsteps= sum(steps)) %>%na.omit()
hist(databydate$tsteps, xlab = "Total Daily Steps",main="Histogram of Total Number of Steps Taken Each Day", breaks = 20)
```  

#### 3. Calculate and report the **mean** and **median** of the total number of steps taken per day  
#### Mean
```{r}
mean(databydate$tsteps)
```  
#### Median
```{r}
median(databydate$tsteps)  
```  
### What is the average daily activity pattern?  
#### 1. Make a time series plot (i.e., type = 1) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)  
```{r}
library(ggplot2)
databyinterval <- stepdata%>% select(interval, steps) %>% na.omit() %>% group_by(interval) %>% summarize(tsteps= mean(steps)) 
ggplot(databyinterval, aes(x=interval, y=tsteps))+ geom_line()
```  

#### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?  

```{r}
databyinterval[which(databyinterval$tsteps== max(databyinterval$tsteps)),]
```
### Imputing Missing Values  
#### 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)  
```{r}
missingVals <- sum(is.na(data))
missingVals
```
#### 2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc. **The startegy to be used for this project to fill in the missing values in the dataset is based on using the mean of the 5-minute interval**  
#### 3.Create a new dataset that is equal to the original dataset but with the missing data filled in  
```{r}
library(magrittr)
library(dplyr)

replacewithmean <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))
meandata <- stepdata%>% group_by(interval) %>% mutate(steps= replacewithmean(steps))
head(meandata)
```  

```{r}
FillSummedDataByDay <- aggregate(meandata$steps, by=list(meandata$date), sum)

names(FillSummedDataByDay)[1] ="date"
names(FillSummedDataByDay)[2] ="totalnumberofstepsperday"
head(FillSummedDataByDay,15)
```  
#### 4. Make a histogram of the total number of steps taken each day
```{r}
hist(FillSummedDataByDay$totalnumberofstepsperday, xlab = "Steps", ylab = "Frequency", main = "Histogram of Total Number of Steps Per Day", breaks = 20)
```  

#### Calculate and report the mean and median total number of steps taken per day  
```{r}
mean(FillSummedDataByDay$totalnumberofstepsperday)
median(FillSummedDataByDay$totalnumberofstepsperday)
```  
#### Do these values differ from the estimates from the first part of the assignment?  
```{r}
oldmean <- mean(databydate$tsteps, na.rm = TRUE)
newmean <- mean(FillSummedDataByDay$totalnumberofstepsperday)
```  
```{r}
oldmean
```
```{r}
newmean
```  
#### **Based on the mean values returned, *oldmean and newmean*, the values DO NOT differ from the estimates from the first part of the assignment**  

#### What is the impact of imputing missing data on the estimates of the total daily number of steps?  **The impact of imputing missing data on the estimates on the total daily number of steps is NEGLIGIBLE**  
### Are there differences in activity patterns between weekdays and weekends?  
#### Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.  
```{r}
meandata$date <- as.Date(meandata$date)
meandata$weekday <- weekdays(meandata$date)
meandata$weekend <- ifelse(meandata$weekday=="Saturday" | meandata$weekday=="Sunday", "Weekend", "Weekday")
```  
#### Make a panel plot containing a time series plot (i.e. type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.  
```{r}
library(ggplot2)
meandataweekendweekday <- aggregate(meandata$steps , by= list(meandata$weekend, meandata$interval), na.omit(mean))
names(meandataweekendweekday) <- c("weekend", "interval", "steps")

ggplot(meandataweekendweekday, aes(x=interval, y=steps, color=weekend)) + geom_line()+
facet_grid(weekend ~.) + xlab("5-Min Interval") + ylab("average across all weekdays and weekends") +
    ggtitle("Time Series Plot - Average Number of Steps Taken")
```  

